\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-22 jasa/paper-jasa.tex

One strength of this approach is that it factorizes the approximation of the
per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx p(\hat{s}(\mathbf{x})|\theta)$).
Thus, even if the classifier does a poor job at reproducing the level sets of
the per-event likelihood ratio, the density of $\hat{s}$ can still be well
calibrated. In that case, one might loose power, but the resulting inference
will still be valid.\glnote{Shouldn't we elaborate?} This point was made by
\cite{Neal:2007zz} and is well appreciated by the particle physics community
that typically takes a conservative attitude towards the use of machine learning
classifiers precisely due to concerns about the calibration $p$-values in the
face of nuisance parameters associated to the simulator.

@@

We have outlined a technique to reformulate generalized likelihood ratio test
over a high-dimensional data set with multiple events in terms of a univariate
density of a classifier score. We have shown that a parameterized family of
discriminative classifiers $\hat s(x; \theta_0, \theta_1)$ trained and
calibrated with a simulator  $p(x|\theta)$ can be used to approximate the
likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not
possible to directly evaluate the likelihood $p(x|\theta)$.
%A technique for decomposing this ratio when the generative model is a mixture of components was presented with the aim to help focus capacity of the classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small mixture coefficient.  This approach leverages the power of machine learning in a classical statistical setting.
It offers an alternative to approximate Bayesian computation for parameter
inference in the likelihood-free setting that can also be used in the
frequentist formalism without specifying a prior over the parameters. A strength
of this approach is that it separates the quality of the approximation of the
target likelihood from the quality of the calibration. The former is related to
the ability of supervised learning approaches to  classification, which will
continue to improve. The calibration procedure for a particular parameter point
is fairly straight forward since it involves estimating a univariate density
using a generative model of the data. The difficulty of the calibration stage is
performing this calibration continuously in $\theta$. Different strategies to
this calibration are anticipated depending on the dimensionality of $\theta$,
the complexity of the resulting likelihood function, and the difficulties
associated to running the simulator.

>>>>

One strength of this approach is that it factorizes the approximation of the
per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx p(\hat{s}(\mathbf{x})|\theta)$).
Thus, even if the classifier does a poor job at reproducing the level sets of
the per-event likelihood ratio, the density of $\hat{s}$ can still be well
calibrated. In that case, one might loose power, but the resulting inference
will still be valid.\glnote{Shouldn't we elaborate? Since this is one of our selling points, we should give more details in my opinion.} This point was made by
\cite{Neal:2007zz} and is well appreciated by the particle physics community
that typically takes a conservative attitude towards the use of machine learning
classifiers precisely due to concerns about the calibration $p$-values in the
face of nuisance parameters associated to the simulator.

@@

In this work, we have outlined an approach to reformulate generalized likelihood
ratio testing over a high-dimensional data set in terms of a univariate density
of a classifier score. We have shown that a parameterized family of
discriminative classifiers $\hat s(\mathbf{x}; \theta_0, \theta_1)$ trained and
calibrated with a simulator can be used to approximate the likelihood ratio,
even when it is not possible to directly evaluate the likelihood
$p(\mathbf{x}|\theta)$.
% A technique for decomposing this ratio when the generative model is a mixture
% of components was presented with the aim to help focus capacity of the
% classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small
% mixture coefficient.  This approach leverages the power of machine learning in a
% classical statistical setting.
The proposed method offers an alternative to approximate Bayesian computation
for parameter inference in the likelihood-free setting that can also be used in
the frequentist formalism without specifying a prior over the parameters. A
strength of this approach is that it separates the quality of the approximation
of the target likelihood from the quality of the calibration. The former is
related to the ability of supervised learning approaches to  classification,
which will continue to improve. The calibration procedure for a particular
parameter point is fairly straight forward since it involves estimating a
univariate density using a generative model of the data. The difficulty of the
calibration stage is performing this calibration continuously in $\theta$.
Different strategies to this calibration are anticipated depending on the
dimensionality of $\theta$, the complexity of the resulting likelihood function,
and the difficulties associated to running the simulator.
